---
title: "pandas (R)osetta: R"
subtitle: "Intro to R for Pandas users and vice versa."
author: "Lisa Malins"
knit: (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_dir = "output",
      output_format = "all")
    })
output:
  github_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 4
    df_print: "default"
---

&nbsp;

This is a demonstration of basic data wrangling operations in base R.

Sister notebooks demonstrate the exact same operations in Tidyverse R and the Pandas library in Python, just like the Rosetta Stone.

## I/O
### Create dataframe from scratch
```{r}
df <- data.frame(
  letter = c("a", "b", "c", "d", "e"),
  number = c(1:5),
  fruit = c("apple", "banana", "coconut", "date", "elderberry"),
  vegetable = c("arugula", "beet", "carrot", "daikon", "eggplant"),
  name = c("Alice", "Bob", "Carol", "Dan", "Eve")
)
df
```


### Write
```{r}
write.csv(df, "data/R_letters.csv", row.names=FALSE)
write.table(df, "data/R_letters.tsv", sep="\t", row.names=FALSE)
```

### Read
```{r}
read.csv("data/R_letters.csv")
read.delim("data/R_letters.tsv")
```

## Accessing data
Note for Pandas users:

- Both R and Pandas have the same convention of `[row, column]` for retrieving a cell from a dataframe.
- However, if only one number is specified with no comma, in pandas a row is returned, but in R a column is returned.
- Also, remember that **R has 1-based indexing** while Python has 0-based indexing.
- Thus,
  - In pandas: `df.iloc[1]` returns the second row
  - In R: `df[1]` returns the first column

```{r}
# Returns first column ("letter")
df[1]
```

### Select cell
To **select a single cell**, **use double brackets** with the row number, followed by the column number or name.

```{r, results="hold"}
# Get "banana" cell value
df[[2, 3]]
df[[2, "fruit"]]
```

Cells can also be selected with single brackets. The two notations *are* different and are used for different things (more on that later), but when selecting a single value, the distinction is moot in base R.

```{r, results="hold"}
# Usually, single brackets return a dataframe,
# but when it's a 1x1 selection, just the value is returned
df[2, 3]
df[2, "fruit"]
# The 4 expressions above are all identical
```
